\chapter{Beirdd y Tywysogion}
The poetry composed by the \mw{Beirdd y Tywysogion}, or Poets of the Princes, comprises the work of a great variety of poets geographically. These poets lived over a period of several hundred years from eleventh century until the fall of the principality of Gwynedd to the English in 1284.
\section{Unusable examples}

/Gwyn ei fyd bieufai /Gwych ysgor tra m\^or,: lack of syntactic continuity

A byth nis beiddai bai byw: bai after beiddai is two verbs following one another.

CBM	3	26	78	Ef laddai, ef wanai, wanwyd,: wanwyd has no syntactic continuity as wanai and wanwyd are both inflected verbs each forming an independent clause.
\section{Usable examples}
Englynion will be quoted as a full englyn, in other cases, only the line in which \ei\ or \oes\ appears will be quoted. 

\subsection{Meilyr Brydydd}

CBT I: 3: \mw{Mar\w nad Ruffut ab Kynan}

\mwcc{CBT~I:~3,~l.~19}{Gwr a lywei lu kyn bu breua\w d,}{A man led an army before he was destroyed,}
\mwcc{CBT~I:~3,~l.~34}{Rynn ruthrei doruoet oet rybara\w d;}{unyieldingly he would rush upon hosts that were over-eager;}
\mwcc{CBT~I:~3,~l.~36}{Ny vynnei gamhwr garu newba\w d.}{The bender would not love anyone.}
\mwcc{CBT~I:~3,~l.~42}{Pan vei gyfluyt o wyr gwycha\w c,}{When there would be a stir of brave men}
\mwcc{CBT~I:~3,~l.~96}{Gwr a rotei gad kyn dybu y dyt.}{A man would give battle before the day started.}
\mwcc{CBT~I:~3,~l.~102}{Ny bitei diwyth y lwyth osswyt;}{He would not be peaceful towards the enemy hosts;}
\mwcc{CBT~I:~3,~l.~119}{G\w ern gwygid, g\w anei ba\w b yn y gilit,}{A spear was broken, everyone stabbed into each other,}

\subsection{Gwalchmei ap Meilyr}

\mwcc{CBT~I:~6,~l.~17}{Nyd oes gystetlyt y hael heuelyt}{There is no equal to a generous person}
\mwcc{CBT~I:~7,~l.~18}{Ryuei uar\w\ kyn Mada\w c, mad aned!}{He who had been dead before Madog, he is lucky to have been born!}
\mwcc[oesgochel]{CBT~I:~14,~l.~93}{Ac nys oes gochel gochwer\w\ drwyted}{And he has no avoidance of hateful lodging}
Note that example \ref{oesgochel} has a variant reading: \mw{ac nyt gochel gocher\w\ drwyted}
\subsection{Elidir Sais}
\mwcc{CBT~I:~15,~l.~20}{A'm parchai llu meddfaeth.}{And a mead-nourished host would praise me.}

\mwengtr{CBT~I:~17,~ll.~44--47}{No cholled alaf elw freuner --- o'm t\^al,\\Tebygach ym lawer,\\Heb eisieu neb meu o'm ner,\\Oes wogawn hwyldawn haelder}{Instead of loss of wealth, profit of a generous lord --- from my payment,\\It is much more likely to me,\\Without any of me wishing from my lord,\\There is genius-covering glory of generosity}


\mwengtr[es1858]{CBT~I:~18,~ll.~5--8}{Gweithred Tegwared! Cyn ty gweryd --- llawr\\Llawen beirdd o'i wynfyd;\\Gnawd oedd ei wayw yn waedlyd;\\O waedlan, rhuddai wan rhyd.}{The work of Tegwared! Before he had his house on earth --- below\\Poets were happy on account of his delight;\\His spear was frequently bloody;\\From slaughter, the thrust of the ford would redden.}

\subsection{Einion ap Gwalchmai}

\mwcc{CBT~I:~28,~l.~15}{Nyd oes darogan, dyn truan tra\w d,}{There is no prediction, weak wretched man,}

\subsection{Llywelyn Fardd I}

\mwcc{CBT~II:~2,~l.~64}{Yn yt uydei tr\w ch ny bei trechaf,}{Where he who would be wounded if he were not the strongest,}

This example has two data points.

\mwengtr{CBT~II:~12,~ll.~1--4}{Pan \w ei lawen \w rein, pan dy\w rysyei --- waed,\\Pan wyar waryei,\\Pan ryuel, pan rudit e thei,\\Pan Rudlan, pan rudlys losgei,}{When crows would be happy, when blood would flow quickly\\When gore would play,\\When there would be war, when her houses would be reddened\\When Rhuddlan would be there, when the red court would burn,}

\mwengtr{CBT~II:~12,~ll.~5--8}{Pan rudei rudflam, flemychei --- hyt nef\\Yn addef ny noddei,\\Hawd gwelet goleulosc arnei\\O gaer wenn geir amyl Menei.}{When the red flame would redden, it would blaze --- up to the sky\\To be at home would not shelter,\\Easy to see a bright burning on it\\From a white castle at the edge of Menai.}

\subsection{Owain Cyfeiliog}

\mwcc[dwyraigawr]{CBT~II:~14,~l.~1}{Gwa\w r pan d\w yreai ga\w r a dotet,}{The morning, when it would give a shout that must come,}

Example \ref{dwyraigawr} contains emended \mw{d\w yreai}, of original \mw{d\w yre}. Also, \mw{gawr} `shout' is interpreted here by the editors of CBT, rather than \mw{cawr} `giant'.

\subsection{Daniel ap Llosgwrn Mew}

\mwengtr{CBT~II:~18,~ll.~30--33}{A welei Ewein yn lluyt --- yn Lloegyr\\Yn llosci y threuyt,\\Gwelei doryf ar doryf y gilit,\\Gwelei daryf ar dyrrua osswyt.}{Whoever saw Owain in an army in England\\Burning its dwellings,\\He would see a host on a host of his enemy,\\He would see a retreat of the enemy host.}

\subsection{Seisyll Bryffwrch}

\mwcc{CBT~II:~22,~l.~20}{Nid oes pryssur affaith,}{There is no cheerless help,}
\mwcc{CBT~II:~22,~l.~29}{Gwasgarei gweisgfeirch mai mawrthig,}{Strong horses would disperse on martial plains}

\subsection{Gwynfardd Brycheiniog}

\mwcc{CBT~II:~26,~l.~246}{Kyn syrthei urwynen ar urynneu --- o nef,}{Although rushes would fall on hills --- from heaven,}

\subsection{Cynddelw Brydydd Mawr}

\mwcc{CBT~III:~5,~l.~83}{Ac nyd oes vart d\w fyn, dinus gyuyeith,}{And there is no deep bard, skilled in the same language,}
\mwcc[luestaiwynedd]{CBT~III:~8,~l.~65}{Pei byw llary Lleissiawn ni luestai Wynedd}{If Llary Lleission lives, Gwynedd may not camp}
Two data points are found in \ref{luestaiwynedd}
\mwengtr{CBT~III:~11,~ll.~53--56}{Kynnetyf ywch, Bowys, ban wnaeth --- auch gorsaf,\\Auch gorset na bei gaeth,\\Glyw gwyrthua\w r, g\w rthoduch-chwi etwaeth,\\G\w rthodes rywyr, righyllaeth!}{Immunity to you, Powys, when you made a defence,\\Your defence may not be restricted,\\Powerful chief, reject again,\\Strong men withstood, sergeantship! }

\mwcc[Llateillauyn]{CBT~III:~16,~l.~55}{Llatei llauyn gasnar, llathrei galch llassar}{A fierce blade killed, blue enamel shone}

Example \ref{Llateillauyn} contains two data points: first unlenited and the second lenited.

\mwcc{CBT~III:~16,~l.~242}{Yd wanei wanar, yd waned:}{[translation]}
\mwcc{CBT~III:~21,~l.~79}{Kymysgei greu g\w yr a g\w yrt heli --- mor;}{[translation]}
\mwcc{CBT~III:~21,~l.~113}{Branhes ymborthyad, nyd amborthei --- gabyl,}{[translation]}
\mwcc{CBT~III:~21,~l.~162}{Yd gilyei pob llwfyr yny llated,}{[translation]}
Probably a research exception.
\mwcc{CBT~III:~21,~l.~163}{Yd gyrchei pob glew mal yd glywed,}{[translation]}
And so is this one.
\mwcc{CBT~III:~21,~l.~166}{Yd gystlynei ba\w b o'e gystlyned,}{[translation]}
\mwcc{CBT~III:~24,~l.~100}{Eurglet kymynet, kymynei benn:}{[translation]}
\mwcc{CBT~III:~24,~l.~118}{Mygyr yt latei Loegyr hyd lewdir trenn;}{[translation]}
\mwcc{CBT~III:~24,~l.~150}{Ny'm llauarei-y nep na'm bei diga\w n}{[translation]}
\mwengtr{CBT~III:~25,~ll.~9--12}{Kelennic ruteur a'm rotei --- Ririd,\\Ny'm rotes a uei lei,\\Nyd aruan\w l vut uytei,\\Nyd ar uanarch y'n parchei.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~III:~25,~ll.~17--20}{Gwnaeth Goronw, g\w r anlew,\\Gyulafan anhalediw,\\Adlat Ririd, rwyf ana\w,\\A byth nus beitei bei byw!}{[translation\\translation\\translation\\translation]}
\mwcc{CBT~III:~26,~l.~34}{G\w asgarei g\w eiscueirch mei muner}{[translation]}
\mwcc{CBT~III:~26,~l.~92}{O lid, a lithyei gana\w\ llwyd}{[translation]}
\mwengtr{CBT~III:~29,~ll.~17--20}{Bletynt, bleit adwy, yn adwyn yd las,\\Yd lessid a uei uwy,\\Mabolaeth uetuaeth uaccwy,\\Mab o oed a'e hoed yn hwy}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~III:~29,~ll.~21--24}{Hwyaf y'm athreit athreisuar --- y dwyn,\\Ha\w l wenwyn hwyl wanar\\Yn y redei rut wyar,\\Raeadyr ar paladyr ar par.}{[translation\\translation\\translation\\translation]}
\mwcc{CBT~IV:~1,~l.~26}{Ar edrywet gwaed gwaetei gigurein:}{[translation]}
\mwcc{CBT~IV:~3,~l.~29}{Krynei Uaes Carnet rac carnnyal y ueirch,}{[translation]}
\mwcc{CBT~IV:~3,~l.~31}{Krynei ulaen bragad, brad gad greula\w n,}{[translation]}
\mwcc{CBT~IV:~3,~l.~33}{Krynei elorueirch bre[~] bryd na\w n,}{[translation]}
\mwcc{CBT~IV:~3,~l.~34}{Crynei eleurut rac y we[~],}{[translation]}
\mwcc{CBT~IV:~3,~l.~35}{Crynei gylch cadlan, cadle estra\w n,}{[translation]}
\mwcc{CBT~IV:~3,~l.~36}{Crynei galch llassar, llas Eithinya\w n!}{[translation]}
\mwcc{CBT~IV:~3,~l.~37}{Crynei [~]yf a ch\w r ysgwyda\w r}{[translation]}
\mwcc{CBT~IV:~3,~l.~39}{Crynei ys[~]n is Gwennfynna\w n,}{[translation]}
\mwcc{CBT~IV:~4,~l.~62}{Aer oruc oreu bei canmlwyt,}{[translation]}
\mwcc{CBT~IV:~4,~l.~103}{Uch gwaetlyry gwaedlan\w\ g\w aetei gigurein,}{[translation]}
\mwcc{CBT~IV:~4,~l.~130}{Oet trama\w r tremynei tachwet,}{[translation]}
\mwcc{CBT~IV:~4,~l.~209}{Ystre hynt wastad, g\w estei --- gwynuydic,}{[translation]}
\mwengtr{CBT~IV:~5,~ll.~69--72}{Haetassei wa\w r glyw, glod degyn --- gyuet,\\Gwira\w d uet, gwir oet ynn;\\G\w ae vi am uab Gennillyn,\\G\w ae ueirt Kymry g\w edy Gwynn!}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~IV:~6,~ll.~46--49}{Kyngen gywrysset am gywrys --- Deheu,\\Llary Lacheu Heuelys;\\Llathrei lafyn o'e la\w\ ar gyuys,\\Llachar glaer yn aer, yn dyrys;}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~IV:~6,~ll.~102--105}{Gwasga\w d gwa\w d, gwaet nenn, g\w r yn gwisga\w\ --- penn,\\Gwasgarei ba\w b racda\w;\\Gosgort \w rt a ordyfyn preitya\w\\Gosgrynwent cadwent ked walla\w.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~IV:~6,~ll.~226--229}{Ymdibustlei lew ar lan Aeron --- berth\\Pan borthes eryron.\\Rydirwy dyra\w r y hollyon,\\Rydyla\w ch eirchyad ac eirchon,}{[translation\\translation\\translation\\translation]}
\mwcc{CBT~IV:~7,~l.~14}{Llunyei glod glew dihauarch,}{[translation]}
\mwcc{CBT~IV:~9,~l.~34}{A'e gweles, gwelei ryueta\w d:}{[translation]}
\mwcc{CBT~IV:~9,~l.~97}{Llathrei lafyn uch auyn, uch afwyn,}{[translation]}
\mwcc{CBT~IV:~9,~l.~99}{Llatei dreic ar dragon drylwyn}{[translation]}
\mwengtr{CBT~IV:~12,~ll.~9--12}{Ked bei teu, wledic, hyd wlad bor --- Eura\w c,\\Aeruarcha\w c, deifnya\w c dor,\\Rybytei ueu ueith ragor,\\Rybyt\w n bennkert benn-cor.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~IV:~12,~ll.~17--20}{Keb bei teu, wledic, wlad Run --- am Maelg\w n,\\Maelgynig aborthun,\\Rybytei ueu uet annhun,\\Rybyt\w n bennkert, benn-kun.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~IV:~13,~l.~18}{Cadwent pan gyrchaist, gyrch union --- sengi,\\Sangei bawb ei alon;\\Parthred gwyth mal Gwaith Faddon,\\Porthathwy, myrdd ofwy Mon.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~IV:~14,~ll.~9--12}{Cerddawr, cyngraffawr cyngreffin --- ar bryf\\O bryfed cyntefin,\\Barddai ceir, beiddiai byddin,\\Bangawr cyw, llyw lledfegin.}{[translation\\translation\\translation\\translation]}
\subsection{`Prydydd y Moch'}
\mwcc{CBT~V:~1,~l.~12}{Preityei ba\w b y bob digarad;}{[translation]}
\mwengtr{CBT~V:~10,~ll.~89--92}{Hwnn a duc ragor ragoch, --- wyr Kymry,\\Kymriwei dur uochuoch\\Pan \w u lew Loegrwys uoloch,\\Pan oet llif Keirya\w c lliw coch.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~V:~13,~ll.~9--12}{Keda\w l, arwyna\w l arwyneb --- Prydein,\\Prydytyon dioheb,\\Ny uagei uyg\w l atteb,\\Ny uaeth negyuaeth y neb.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~V:~13,~ll.~13--16}{Naf eryr, teyrn rac tympestyl --- kedeyrn,\\Cadwei dud y dan destyl,\\Ner trymder tromgad diwestyl,\\Neu'n gwneir uegys Gweir uab G\w estyl.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~V:~16,~ll.~17--20}{Gruffut grym euruut \dots\ adwyn, --- lew\\Yn lle bei gyuamwyn,\\Unbyn detyf, unboen o'e ddwyn,\\Enbyd y'm ergyd ma\w rgwyn.}{[translation\\translation\\translation\\translation]}
\subsection{Phylip Brydydd}
\mwcc{CBT~VI:~14,~l.~11}{Moladwy uleidriw ny'm rodey brit;}{[translation]}
\mwcc{CBT~VI:~14,~l.~25}{A bei Gwiawn o'e dawn a dadenyt}{[translation]}
\mwcc{CBT~VI:~14,~l.~30}{Tra \w ei dywawl Duw yn y dwy\w dit}{[translation]}
\mwcc{CBT~VI:~14,~l.~54}{Pan \w ei parabl doeth a dywettit,}{[translation]}
\mwcc{CBT~VI:~14,~l.~56}{Y'm hadawssey Rys na'm diddolit.}{[translation]}
\subsection{Dienw 4}
\mwengtr{CBT~VI:~20,~ll.~37--40}{Pebyllywys vy lly\w\ yn llu hyfryt --- praff,\\A'r pryff\w n o Wyndyt,\\Gorthoei drei dra\w s a hyt,\\Gorthir y gelwir Geirwryt.}{[translation\\translation\\translation\\translation]}
\subsection{Dafydd Benfras}
\mwcc{CBT~VI:~25,~l.~37}{Ni allai Daliessin.}{[translation]}
\mwcc{CBT~VI:~27,~l.~14}{Cyn bai trwch, na bai trychanmlwydd.}{[translation]}
\mwcc{CBT~VI:~27,~l.~40}{Ni gaffei llew gryd greidiaw awydd,}{[translation]}
\mwcc{CBT~VI:~27,~l.~81}{Am drais Duw, nid oes dim gyhyd,}{[translation]}
\mwengtr{CBT~VI:~32,~ll.~5--9}{Llynguawr daear awr deruyn, --- a garaf,\\A gereis y chanlyn;\\Neut eddwyf yn oer dyddyn,\\Nyt oes dim einoes y ddyn.}{[translation\\translation\\translation\\translation]}
\mwcc{CBT~VI:~33,~l.~63}{Bydei praff hedi\w\ y prynet --- o ved}{[translation]}
\subsection{Dienw 5}
\mwcc{CBT~VI:~36,~l.~15}{Kynn bei Mab y'n Tat, rat r\w y dannwed,}{[translation]}
\subsection{Prydydd Bychan}
\mwengtr{CBT~VII:~2,~ll.~13-16}{Bytinoet hart oet dyt y hyrtei --- Rys\\Ki\w da\w d Ros ohanei,\\Lleith kynnygyn, kynnetyf G\w alchmei,\\Lle\w\ ga\w r ysg\w ydua\w r, ys gwnai.}{[translation\\translation\\translation\\translation]}
\mwengtr{CBT~VII:~7,~ll.~9--12}{Creir nyd amharchei creula\w n --- y volchlafyn\\A'e ualchlys yg Cara\w n,\\Rwyf blwng, r\w yt yst\w ng estra\w n,\\Rot digraf, rotyo digra\w n.}{[translation\\translation\\translation\\translation]}
\mwcc[parcheiparch]{CBT~VII:~15,~l.~15}{G\w r a'm parchei, parch dilesc,}{[translation]}
I wonder if \ref{parcheiparch} is a usable example.
\mwcc{CBT~VII:~15,~l.~16}{Ac a'm rotei rut eurwisc.}{[translation]}
\mwcc{CBT~VII:~18,~l.~21}{Kynnatyl, lle bei beilch ddoethon}{[translation]}
\mwcc{CBT~VII:~18,~l.~25}{A grettei Wen, walch syber\w,}{[translation]}
\subsection{Hywel Foel}
\mwcc{CBT~VII:~23,~l.~10}{G\w r, pei ryt, ual Run uab Beli,}{[translation]}
\mwcc{CBT~VII:~23,~l.~11}{G\w r ny adei Loegyr y losgi --- y deruyn.}{[translation]}
\subsection{Llygad G\^wr}
\mwcc[oesrwyf]{CBT~VII:~24,~l.~38}{Nyd oes, rwyf eryoes aer dyuysgi.}{[translation]}
Example \ref{oesrwyf} is very similar to \ref{parcheiparch}
\mwcc{CBT~VII:~24,~l.~97}{Gwelei wa\w r ar wyr lluossyt}{Saw a chief of hosts of men}
\mwcc{CBT~VII:~24,~l.~99}{A welei Lewelyn, lewenyt --- dragon,}{That saw Llewelyn, rejoicing --- hero}
\mwcc{CBT~VII:~25,~l.~9}{Pan uei urwydyr, eurgrwydyr, aergun dyged,}{When there would be a fight, gold-embossed, a battle-lord of fate,}
\subsection{Iorwerth Fychan}
\mwengtr[beiron]{CBT~VII:~31,~ll.~1--4}{Pei kawn o gyflwr gyfle broui --- rin,\\Kyt bei ron vyng krogi,\\Vy neges oed vynegi\\Vyn gouec, dyn tec, y ti.}{If we have in this state a chance to experience --- intimacy\\Even if I were hanged \\My message would be to proclaim\\My affection, pretty one, to you.}

Although lenition of \mw{r} is not written in the Middle Welsh period, the \textit{cynghanedd} makes a suggestion as to lenition of \mw{ron}: \mw{ron} alliterates with \mw{rin} `intimacy' here, making a \textit{cynghanedd draws}. There is no reason for \mw{rin} to be lenited here, so non-lenition of \mw{ron} would make for better alliteration. However, maybe this example must be considered a research exception, as \mw{bei ron} may be considered one word \parencite[s.v.\ \textit{perh么n, peirh么n, pei rh么n, pe rh么n, \&c.}]{bevan_geiriadur_2014}. This would make lenition or lack of lenition dependant on internal sandhi rules rather lenition following \ei.
\subsection{Gruffudd ab yr Ynad Coch}
\mwcc{CBT~VII:~36,~l.~73}{Nyt oes le y kyrcher rac carchar bra\w,}{There is no place gone to before a frightful prison,}
\mwcc{CBT~VII:~36,~l.~74}{Nyt oes le y trigyer: och o'r trigya\w!}{There is no place stayed at: woe the staying!}
\mwcc{CBT~VII:~37,~l.~1}{Pei byw vy llyw, llew flamdur}{If my boss were alive, lion with a flaming sword}
\mwcc{CBT~VII:~37,~l.~3}{Ny ladei gat goet Aber,}{An army would not kill the forest of Aber,}
\mwcc[chryneibop]{CBT~VII:~38,~l.~7}{Na chrynei bop pryt rac pryder -- c\w ympa\w}{Nor would he tremble everytime out of fear --- of falling}

Example \ref{chryneibop} has a variant reading for \mw{chrynei}, i.e. \mw{\d{c}yfyd}. Unusable anyway, because \mw{bop} is a research exception, and because \mw{bop pryt} is an adverbial clause of time following \mw{chrynei}, and may be lenited for this reason only.
\mwcc[oesvrenhin]{CBT~VII:~41,~l.~10}{Nyt oes, Vrenhin, Deus domin, dimynn eisseu.}{There is not, King, Lord God, any wish.}

The translation \ref{oesvrenhin} suggests that this is a research exception, since \mw{Vrenhin} `king' does not serve as a complement to \mw{oes}. Rather, it is lenited because of its use as a vocative.
\mwcc[oeswadeu]{CBT~VII:~41,~l.~30}{Geyr bronn Douyd a Ssywedyd, nyt oes wadeu;}{In presence of the Lord and teacher, there are no contradictions;}

Lenition of \mw{wadeu} following \oes\ in Example \ref{oeswadeu} is confirmed both on the basis of orthography and \textit{cynghanedd}. It is a \textit{cynghanedd sain}, where \mw{Douyd} `Lord' rhymes with \mw{Ssywedyd} `teacher', which in turn also alliterates with \mw{(s-) wadeu} `contradictions'.

\mwcc[oesallu]{CBT~VII:~41,~l.~31}{Nyt oes allu yn erbyn Iessu, na synn droeu,}{There is no power against Jesus nor dazed wandering}

In Example \ref{oesallu}, \mw{allu} `power'  is lenited following \oes. This may not be confirmed nor disproven by \textit{cynghanedd}.
\mwcc[wnelheivegys]{CBT~VII:~42,~l.~3}{Pei na wnelhei vegys y harchei}{If a man would not do as he commanded}
Example \ref{wnelheivegys} looks like an unusable example, given that \mw{vegys} `as' introduces an adverbial clause rather than serving as a complement to \mw{wnelhei} `would do'.
\mwcc[medylyeidyn]{CBT~VII:~43,~l.~9}{Bei medylyei dyn a'e vedylya\w}{If a man would think with his thinking}

Variant reading in \ref{medylyeidyn}: \mw{Pe me\dd ylie}. 
\subsection{Bleddyn Fardd}
\mwengtr{CBT~VII:~47,~ll.~5--8}{Rwy y roddei wr, a orwu --- e glod\\Ym pob g\w lad y rywu,\\Rut\w eirch (ryt\w eith eu kyrchu);\\Ruyt \w al Nut, rot\w a\w r \w ut \w u.}{Excessively the man would give, which surpassed --- his glory\\In every land in which he was,\\Red horses (A free deed to approach them)\\Noble as Nudd, his favour was generous}
\mwengtr{CBT~VII:~56,~ll.~13--16}{Oet plyc dwyn terrwyn yn y tyrrei --- bobyl\\O bobman a'y kyrchei,\\Oet dily\w yn eur a dalei,\\Oet dilut but myn y bei.}{Humiliation was to bring the fierce who gathered --- people\\Who meet him from everywhere,\\The gold he paid was smooth\\The spoils of where he would be were unreluctant.}
