\chapter{Case studies on the \mw{Hengerdd}}
The aim of this chapter is to demonstrate how orthography of lenition may be used to arrive at the dating of a text and perhaps to identify scribes. I have analysed two sources: the Book of Aneirin and the Book of Taliesin. The former is interesting because it consists mainly of a single tale, the \mw{Gododdin}. Parts of this poem survive in three versions in two hands, offering evidence of divergence. The existence of several copies allows for the application of text-critical methods in order to reconstruct various stages of modernisation, and it is interesting to see how such a reconstruction tallies with the representation of lenition in the respective versions. The Book of Taliesin contrasts with this in several ways: the whole manuscript is written by a single scribe, and it contains a wide variety of poems from a wide variety of periods. The existence of a single scribe allows us to assume a certain consistency in how lenition was modernised, while the variation in source age allows us to reconstruct how lenition was represented in different stages of the Welsh language.
\input{3orth/aneirin}

\section{Conclusion}
It is important to stress that it is not the relative frequency of orthographically represented lenition that may be used to date the period in which an original composition was written, or in which lenition was added to such a composition, but the distribution of orthographic lenition across consonants. 
If lenition is written similarly irregularly for each consonant, it may be thought to be written before orthographic lenition. 
If a text has consistent written lenition of consonants other than voiceless stops, but inconsistent writing of voiceless stops, then we may assume that it was originally written in a time where only the former were represented orthographically. 
If a text writes out lenition in full, then its original composition may be thought to have been composed after lenition of voiceless stops came to be represented.

In the Book of Aneirin, the degree to which lenition is represented does not seem to depend on the scribe. Both scribe A and B reach a represenation rate of some eighty per cent in some parts, while representation of lenition is scarce in the last pages of the manuscript. Rather, it seems that lenition has been added to a common exemplar. This assumption allows us to make sense of how lenition is distributed across the Book of Aneirin. Going further we may put rough dates on when hypothetical manuscripts in a stemma were produced.

In the Book of Taliesin, it seems that irregularly represented lenited consonants\footnote{i.e.\ all consonants in \mw{Trawsganu Kynan}, voiceless stops only in other texts} are consistent in their irregularity. That is, the scribe added lenition with about sixty per cent accuracy. This implies that  relative frequency of orthographically represented lenition is a variable dependent on the scribe modernising, and perhaps each scribe may have his own percentage assigned to him, given a sufficiently large corpus. This may make the relative frequency with which lenition is represented a useful statistic for identifying scribes. 

The sixty per cent mark in the Book of Taliesin contrasts markedly with the eighty per cent mark in the Book of Aneirin: the difference between these percentages means that whoever modernised the representation of lenition in the Book of Aneirin (with the exception of the final pages) missed out on only half as many cases of lenition as the scribe of the Book of Taliesin did. This also confirms that these percentages without considering consonant type do not tell anything about the age of the original composition. A case in point is comparing \mw{Gododdin} A with \mw{Armes Prydein}. Lenition has a representation rate of 82\% in the former and 76\% in the latter. Considering how lenition is an innovation, one might be tempted to think that the text with comparatively fewer instances of lenition should be the older one. However, few people would argue that \mw{Armes Prydein} is more archaic than the \mw{Gododdin}, let alone on orthographical grounds.

It is theoretically possible that a scribe could reach one hundred percent accuracy in updating the representation of lenition. This would mean that there would be no practical difference between such a modernised text and a text composed after lenition came to be represented orthographically in full.