\chapter{Lenition after \ei\ and \oes\ and parallel cases}
\label{postverballenition}
Following \ei\ and \oes, lenition is not typically written for voiceless stops in early texts, even though other consonants are lenited following these verbs. This type of lenition is not a typical orthographical aberration of this manuscript, however. Although the White Book recension of \mw{Culhwch ac Olwen} does not write lenition with 100\% accuracy, failure to represent lenition orthographically is not typically dependent on consonant type. Compare, for example, all instances of \mw{deu, dwy} `2' in the White Book recension of \mw{Culhwch ac Olwen}. As can be seen in Table \ref{deudwy}, only two examples fail to lenite, and these are one voiceless stop and one other type of consonant\footnote{Moreover, the only example suggesting non-lenition, \mw{deu par} is problematic, because \gmw[two pairs]{deupar} is also attested as one word~\parencite[deubar, deupar]{bevan_geiriadur_2014}.}.

\begin{table}[h]
\centering

\begin{tabular}{@{}lllll@{}}
\toprule
\textbf{\textbf{column}} & \textbf{\textbf{line}} & \textbf{\textbf{gloss}} & \textbf{\textbf{lenited}} & \textbf{\textbf{\mw{p, t, c}}} \\ \midrule
455 & 1 & \mw{deu par} & no & yes \\
455 & 13 & \mw{deu uilgi} & yes & no \\
455 & 21 & \mw{dỽẏ morwennaỽl} & no & no \\
467 & 17 & \mw{deu geneu} & yes & yes \\
467 & 35 & \mw{deu was} & yes & no \\
468 & 12 & \mw{deu was} & yes & no \\
469 & 13 & \mw{dỽẏ goẏs} & yes & yes \\
478 & 5 & \mw{dỽẏ garant} & yes & yes \\
480 & 27 & \mw{deu gẏtbreinaỽc} & yes & yes \\
480 & 33 & \mw{deu gẏtbreinhaỽc} & yes & yes \\
484 & 9 & \mw{deu geneu} & yes & yes \\
484 & 17 & \mw{deu geneu} & yes & yes \\ \bottomrule
\end{tabular}

\caption{Lenition after \mw{deu, dwy} `2' in the \gls{wbr} recension of \mw{Culhwch ac Olwen}}
\label{deudwy}
\end{table}

I hypothesise that lenition following \ei\ and \oes\ is a remnant of an older Welsh phonology where voiceless stops could lenite, but where a lenited voiceless stop did not equal a radical voiced stop word-initially. Because of this <b, d, g> could not be used to denote the lenited versions of these voiceless stops. Consequently, lenition was not written for these consonants. Later scribes modernising the orthography of  \mw{Culhwch ac Olwen} and \mw{Pwyll Pendeuic Dyuet} generally managed to write lenition of voiceless stops, but failed to do so following these specific verbal endings.

This begs the question what makes the lenition following \ei\ and \oes\ so special. Why did scribes fail to modernise lenition specifically in these cases, and are these cases unique? It must be noted that contact lenition following these verbs remained productive up until much later in the history of the Welsh language. Contact lenition following \ei\  remained a feature of Welsh even after object lenition had become fully productive~\autocite[42]{van_sluis_development_2014}. Lenition following \oes\ also remained, and even analogically spread to other verbal forms used as existential verbs~\autocite[29, 39]{van_sluis_development_2014}.

Nevertheless, lenition following these verbs has no grammatical function. Unlike lenition following \mow[his]{ei}, for example, lenition following these verbal endings has no bearing on the resulting semantics of the utterance. Lenition following these verbs also differs from e.g.\ lenition following feminine nouns in that the lenited word does not necessarily agree grammatically to the word causing lenition. Instead, postverbal lenition may even operate across constituent boundaries between the VP and the NP, \eg when the subject follows the verb. These properties make postverbal lenition a relatively opaque type of lenition\todo{`Opaque' may not be the most descriptive term for what I mean here.}.

It stands to reason that this `opacity' was not unique to postverbal lenition, and perhaps similarly opaque types of lenition were also not modernised. 

\section{Nouns in apposition}
\textcite[\S19]{evans_grammar_1964} notes that nouns in apposition to a personal name are lenited, but gives two exceptions to this rule: \mw{Pwyll \textbf{P}endeuic Dyuet} `Pwyll, Prince of Dyfed', and \mw{Pwyll \textbf{P}en Annwn} `Pwyll, Lord of Annwn'. Both cases show a non-lenited \mw{p} standing in apposition. However, Evans also lists \mw{Rita \textbf{G}awr} `Rita the Giant' among the positive examples. This matter deserves a more thorough investigation.


\input{3orth/apposition}