\chapter{Scribe X86}
The scribe of the Book of Taliesin, of whom we now know that he modernised some sixty percent of lenitions in this book, has been the scribe of several other works. This figure of sixty percent modernisation is fairly conservative, so any other text copied by this scribe may show similarly old patterns leaking through a partially updated Middle Welsh text. This scribe is often known as the  'Book of Taliesin Scribe' and has been given the designation `X86' by Daniel Huws\todo{where?}. Some aspects of his orthography imply that that the scribe was associated with mid- or south-east Wales. These are writing Standard and Northern Welsh \mw{chw} with \mw{wh} or \mw{hu}, and avoidance of \mw{\ci} in words such as \mw{keinhawc}~\autocite[2]{haycock_legendary_2015}.  

Haycock gives the following list of works written by the same hand~\autocite[2]{haycock_legendary_2015}. Some of these texts, such as \mw{Llyfr Cyfnerth}, also betray a connection with South Wales in terms of content:

\begin{itemize}
    \item \gls{nlw} Mostyn 117 (\mw{Brut Dingestow})
    \item \gls{nlw} Peniarth 2 (\mw{Llyfr Taliesin})
    \item \gls{nlw} Peniarth 6, part iv (Part of \mw{Geraint ac Enid})
    \item \gls{bl} Harley 4353, also known as Manuscript V (A law book from the \mw{Cyfnerth} type)
    \item \gls{bl} Cotton Cleopatra A.xiv (idem)
\end{itemize}

It is interesting to see how these texts modernise lenition. In order for these texts to be able to show this, the urtext behind these recensions must be older than when lenited voiceless stops came to be represented. Otherwise, there would be nothing to modernise in this respect. \gls{nlw} Mostyn 117 is a copy of \gls{nlw} 5266, which in turn dates from the late thirteenth century. By this period, lenited voiceless stops were already represented, so this text should not be useful for analyzing the development of lenition. \gls{nlw} 5266 shows lenition of voiceless stops in the majority of cases. As such, the percentage of represented voiceless stops in \gls{nlw} Mostyn 117 does not tell us much about the behaviour of scribe X86. However, one could devise an alternative method: collect all instances in which lenition fails to be represented orthographically in \gls{nlw} 5266, and see how these are represented in \gls{nlw} Mostyn 117.

The piece from \mw{Geraint} found in \gls{nlw} Peniarth 6 may go back far enough to precede orthographic lenition of voiceless stops. \todo{Look up some literature on the dating of this tale.} This tale's original composition in Welsh is quite a bit younger than some of the prophecies found in the Book of Taliesin, but may still be just about old enough to precede orthographic lenition of voiceless stops. The tale's relative recentness may in fact be an advantage, as it may tell us about how and where lenition of voiceless stops first came to be written. After all, this innovation surely did not suddenly spring out of nothing throughout Wales. It seems that \mw{Geraint} is based on the French tale \textit{Erec et Enide} by Chrétien de Troyes, but the exact relationship is still debated. Possibly, an original French composition predated the Welsh, but the Welsh predated the version by Chrétien de Troyes. The twelfth century seems a good possibility for the original Welsh composition\todo{Look up Arthur of the Welsh}. A text composed in this period would be expected to have orthographical lenition, but not of voiceless stops.

The law texts look promising. Both texts offer a balanced mix of represented and non-represented lenition at first blush. Moreover, the older law texts of \mw{Llyfr Iorwerth} in the Black Book of Chirk show practically no lenition of voiceless stops at all. The law texts should therefore provide a good impression of what scribe X86 did in terms of modernising lenition.

All in all, the material of the Book of Taliesin combined with the other work done by this scribe gives a varied corpus in genre and date of original composition. Analysis of his work may be expected to stretch the limits of the proposition that a scribe may be identified by the rate at which he modernises lenition. Alternatively, this varied corpus may allow us to see how a scribe may have modernised lenition at different rates depending on genre, date of composition or other linguistic features found in it.

% \section{NLW 5266 towards Mostyn 117}
% NLW 5266 does show occasional representation of lenited voiceless stops, but not  completely. Consider the following lines from NLW 5266:

% \begin{mwl}
% \item%
%     \mw{%
%     Ac y \al{gyd} a hynny kyn yd oet dec ar arderchaỽoc. o ỽyth dynas ar ugeynt o \al{pryf} dynassoed a rei o hynny ysyd hedyỽ yn dyffeyth gỽedy eu gỽasgaru ac eu dystryw a gadu gỽall arnadunt ereyll etwan ysyd yn seuyll yn \al{gyỽan} ac yn \al{gyweir} o wuroed cadarn goruchel arderchaỽc. 
%     }
%     (NLW 5266 p.~2, ll.~16--21)
% \end{mwl}

\input{3orth/taliesin}


\section{Harley 4353}
Let's take a look at one of the law manuscripts scribe X86 wrote: \gls{bl} Harley 4353. This manuscript contains a recension of \mw{Llyfr Cyfnerth}. The Black Book of Chirk, a member of the Iorwerth branch of the law's of Hywel Dda already shows that voiceless stops do not generally have their lenition written. It is therefore similarly reasonable to assume that an exemplar of \gls{bl} Harley 4353 did not have lenition of voiceless stops written out. Whether the manuscript exemplar before X86's eyes did not have represented lenition of voiceless stops is harder to ascertain. If the patterns found in this manuscript are very similar to what is found in the Book of Taliesin, then it would seem obvious that it is the very same scribe who added them. 

Copying mistakes may also give grounds to assume that an exemplar did not write lenition of voiceless stops. Consider the following excerpt (translation by Wade-Evans):
\mwcc{Harley 4353, f.~8v, ll.~12--13}{Or a y penkynyd yn anreith gan y teulu y brenhin.}{If the chief huntsman goes to foray with the king's household [\dots]}
In the phrase \mw{gan y teulu y brenhin} `with the king's household', the general meaning is well understood, but the first \mw{y} here does not quite make sense. In such a genitive construction, the first article must be deleted. Interpreting \mw{y} as a possessive pronoun does not make any more syntactic or semantic sense, as the household is the king's; not the chief huntsman's. I assume that scribe X86 wrote \mw{gan y} where he should have written \mw{y gan}, a compound preposition also meaning `with'. This compound preposition is frequently found elsewhere in this text. The effect of this error is that the lenition of \mw{teulu} `household' could not be added following \mw{gan}, because it no longer immediately followed \mw{gan}.

Table~\ref{lenitionharley4353} gives an overview of how lenition is represented in this manuscript up to f.\ 8v. The results are found in table~\ref{overviewharley4353}. We find that 68.1\% of lenited voiceless stops has lenition represented orthographically. The amount of non-represented other consonants is negligible. This figure is slightly higher than what is usually found in the Book of Taliesin, but still lower than what is found in the Book of Aneirin. 

\begin{table}[h]
\centering
\begin{tabular}{@{}rlll@{}}
\toprule
 & \textbf{Rep.} & \textbf{Not rep.} & \textbf{Total} \\ \midrule
T & 139 & 65 & 204 \\
Not T & 101 & 6 & 107 \\
\textbf{Total} & \textbf{240} & \textbf{71} & \textbf{311} \\ \bottomrule
\end{tabular}
\caption{Correspondence between initial consonant type and orthographic representation of lenition in the Harley 4353  recension of \mw{Llyfr Cyfnerth} up to f.\ 8v.}
\label{overviewharley4353}
\end{table}

\section{\mw{Geraint} from Peniarth 6}

\begin{table}[h]
\centering
\begin{tabular}{@{}rlll@{}}
\toprule
 & \textbf{Rep.} & \textbf{Not rep.} & \textbf{Total} \\ \midrule
T & 33 & 21 & 54 \\
Not T & 80 & 2 & 82 \\
\textbf{Total} & \textbf{113} & \textbf{23} & \textbf{136} \\ \bottomrule
\end{tabular}
\caption{Correspondence between initial consonant type and orthographic representation of lenition in pp.\ 17--19 of \mw{Geraint ac Enid} in \gls{nlw} MS.\ Peniarth 6.}
\label{overviewgeraint}
\end{table}

Table \ref{overviewgeraint} shows that lenition of voiceless stops is represented in 61.1\% of instances, and that non-representation of other consonants is again negligible. This percentage corresponds with what is found in the Book of Taliesin. 

This observation makes it likely that scribe X86 was the one who added orthographic representation of lenited voiceless stops to Peniarth 6. This obviously means that his exemplar did not have written lenition in these instances. This has implication for the date of original composition in Wales of \mw{Geraint}. \todo{I would not mind taking a closer look at WBR and RBH here}



\section{\mw{p} vs.\ \mw{t} vs.\ \mw{c}}
At this stage, it is appropriate to unravel further how the sixty per cent mark is formed. It is unlikely that the scribe threw dice in order to decide where to represent lenition and where not to. Rather, we must look for a way to make sense of this distribution. Distinguishing between the three voiceless stops already goes a long way, as is shown by Table~\ref{perclenptcx86}\footnote{This table does not include \mw{Trawsganu Kynan}, because it stems from a wholly different orthographical tradition, but perhaps that is precisely a reason it should be included, albeit in a separate table perhaps.}.


\begin{table}[h]
\centering
\begin{tabular}{@{}llllll@{}}
\toprule
\textbf{} & \textbf{\textbf{}} & \textbf{\textbf{Harley 4353}} & \textbf{\textbf{Peniarth 6}} & \textbf{AP \& Proph.} & \textbf{\textbf{Total (\%)}} \\ \midrule
\multirow{2}{*}{p} & rep. & 13 & 1 & 14 & \multirow{2}{*}{32.9} \\
 & not rep. & 24 & 10 & 23 &  \\
\multirow{2}{*}{t} & rep. & 5 & 1 & 9 & \multirow{2}{*}{16.9} \\
 & not rep. & 36 & 10 & 28 &  \\
\multirow{2}{*}{c} & rep. & 121 & 5 & 94 & \multirow{2}{*}{92.5} \\
 & not rep. & 31 & 1 & 14 &  \\ \bottomrule
\end{tabular}
\caption{Representation of lenition of different voiceless stops in three manuscripts written by scribe X86.}
\label{perclenptcx86}
\end{table}

The pattern found in this table is quite clear: all manuscripts presumably copied by scribe X86 from an exemplar without orthographical lenition of voiceless stops achieves a near-perfect consistency in representing lenition of \mw{c}, but \mw{p} and \mw{t} barely get represented at all. Haycock already noticed this: `Lenition of initial p, t (and d) are not generally realized' \autocite[p.~7, n.~18]{haycock_legendary_2015}



One question that arises from these tendencies is how the exceptions may be accounted for. Let's take a look at some exceptional cases, i.e.\ non-represented lenition of \mw{c}, and represented lenition of \mw{p} and \mw{t}. In the prophecies we find the following pattern. Mostly petrified and hypercorrect lenition for represented lenited \mw{p, t}. Non-represented \mw{c} has a fair bit of object lenition, which might not be appropriate in the first place. These statistics indeed suggest as much. In \mw{Armes Prydein}, we see that syntactically based mutations are fairly prevalent among represented lenited \mw{p, t}.  The same is the case for unrepresented \mw{c}. In the law text, represented lenited \mw{p, t} are mostly found in the words \mw{bieu} `owns', and where lenition is petrified. Syntactically based lenition is again responsible for many instances of non-represented lenition of \mw{c}. In \mw{Geraint}, the numbers are too small to make any inferences.

Another follow-up question is whether this tendency to write lenition of \mw{c}, but not the other voiceless stops is a particularity of scribe X86, or an intermediate stage in the development of a Middle Welsh orthography that could represent these lenitions. 

A third question is how to make linguistic sense of this pattern. What makes \mw{c} so different from \mw{p} and \mw{t} in the Middle Welsh orthographical system? The Latin alphabet is known to be woefully inadequate for representing the full lineup of Welsh dentals, but the same is not the case with labials and velars~\autocite{russell_rowynniauc_2003}. What we find, however, is that labials and dentals are on one side of the spectrum with velars on the other side, so this line of reasoning does not account for the facts, or at least not on its own.  

Another issue that is applicable to \mw{t}, but not \mw{p} is the matter of South Welsh spelling. As we know, scribe X86 was connected to South Wales. The South Welsh norm was to use \mw{d} for final -/d/, and \mw{t} for final -/ð/. However, Haycock notes that the scribe's norm was as follows: `\textit{t} for final -d; [\dots] \textit{d} for -ð and medial -ð-, and \textit{t} as well as \textit{d} for medial -d-.' \autocite[p.~7, n.~18]{haycock_legendary_2015}. 




\section{The tables}
\begin{mylongtable}{@{}rrlllll@{}}
\toprule
\textbf{Folio} & \textbf{Line} & \textbf{Word} & \textbf{Translation} & \textbf{Cause of lenition} & \textbf{Represented} & \textbf{Starts w/} \\ \midrule
1r & 1 & \mw{brenhin} & `king' & apposition & no & b \\
1r & 1 & \mw{mab} & `son' & apposition & no & m \\
1r & 2 & \mw{wnaeth} & `did' & \mw{a} & yes & g \\
1r & 9 & \mw{paỽb} & `everybody' & \mw{ar} & no & p \\
1r & 10 & \mw{kyfreitheu} & `laws' & compound & no & c \\
1r & 11 & \mw{pop} & `every' & \mw{o} & no & p \\
1r & 13 & \mw{perchen} & `owns' & \mw{o} & no & p \\
1r & 13 & \mw{taf} & (river name) & \mw{ar} & no & t \\
1r & 17 & \mw{wneuthur} & `do' & \mw{y} `to' & yes & g \\
1r & 20 & \mw{wnaethant} & `did' & \mw{a} & yes & g \\
1r & 21 & \mw{wneuthur} & `do' & \mw{-fu} & yes & g \\
1r & 22 & \mw{gynylleitua} & `assembly' & \mw{y} `his' & yes & c \\
1r & 23 & \mw{benbaladyr} & `entirely' & feminine noun & yes & p \\
1r & 23 & \mw{gymry} & `Wales' & feminine \mw{un} & yes & c \\
1r & 25 & \mw{cyfreitheu} & `laws' & \mw{o} & no & c \\
1r & 25 & \mw{penhaf} & `highest' & adverbial clause & no & p \\
1r & 23--24 & \mw{tor/hei} & `breaks' & \mw{a} & no & t \\
1r & 3--4 & \mw{ky/mry} & `Welshmen' & parenthesis & no & c \\
1v & 1 & \mw{vren/hines} & `queen' & feminine article & yes & b \\
1v & 13 & \mw{gan} & `with, by' & petrification & yes & c \\
1v & 13 & \mw{wisc} & `clothing' & preposed adjective & yes & g \\
1v & 14 & \mw{pop} & `every' & adverbial clause & no & p \\
1v & 14 & \mw{gan} & `with, by' & petrification & yes & c \\
1v & 14 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
1v & 16 & \mw{geiff} & `has' & \mw{a} & yes & c \\
1v & 16 & \mw{wlat} & `country' & \mw{e} `his' & yes & g \\
1v & 17 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
1v & 17 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
1v & 19 & \mw{wna} & `does' & \mw{a} & yes & g \\
1v & 20 & \mw{wreic} & `wife' & \mw{y} `his' & yes & g \\
1v & 21 & \mw{ỽr} & `man' & \mw{y} `his' & yes & g \\
1v & 21 & \mw{ỽyd} & `presence' & \mw{y} `his' & yes & g \\
1v & 21 & \mw{latho} & `kills' & \mw{a} & yes & ll \\
1v & 22 & \mw{vo} & `is' & \mw{pan} & yes & b \\
1v & 23 & \mw{telir} & `is paid' & \mw{a} & no & t \\
1v & 25 & \mw{teyrnas} & `kingdom' & \mw{e} `his' & no & t \\
1v & 25 & \mw{gyrhaetho} & `reaches' & \mw{a} & yes & c \\
1v & 17--18 & \mw{gaf/fan} & `receive' & \mw{a} & yes & c \\
2r & 2 & \mw{aran vys} & `ring finger' & \mw{e} `his' & yes & g \\
2r & 2 & \mw{gadeir} & `chair' & \mw{y} `his' & yes & c \\
2r & 3 & \mw{deni} & `under her' & \mw{o} & yes & t \\
2r & 3 & \mw{wyalen} & `rod' & feminine article & yes & g \\
2r & 4 & \mw{anho} & `holds' & \mw{a} & yes & g \\
2r & 8 & \mw{tecceir} & `is upheld' & \mw{a} & no & t \\
2r & 8 & \mw{warthec} & `cattle' & \mw{o} & yes & g \\
2r & 9 & \mw{loscỽrn} & `tail' & \mw{ỽrth} & yes & ll \\
2r & 12 & \mw{telir} & `is paid' & \mw{a} & no & t \\
2r & 13 & \mw{tri} & `three' & \mw{o} & no & t \\
2r & 13 & \mw{gan} & `with, by' & petrification & yes & c \\
2r & 14 & \mw{torher} & `is broken' & \mw{pan} & no & t \\
2r & 14 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
2r & 15 & \mw{pan} & `when' & \mw{neu} & no & p \\
2r & 15 & \mw{traỽher} & `is struck' & \mw{pan} & no & t \\
2r & 15 & \mw{tynher} & `is taken' & \mw{pan} & no & t \\
2r & 15 & \mw{lit} & `anger' & \mw{trỽy} & yes & ll \\
2r & 16 & \mw{treis} & `violence' & \mw{gan} & no & t \\
2r & 16 & \mw{gan} & `with, by' & petrification & yes & c \\
2r & 17 & \mw{telir} & `is paid' & \mw{a} & no & t \\
2r & 17 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
2r & 19 & \mw{veirch} & `horses' & \mw{ar} & yes & m \\
2r & 19 & \mw{wetha} & `submits' & \mw{a} & yes & g \\
2r & 20 & \mw{getymdeithas} & `company' & \mw{y} `his' & yes & c \\
2r & 21 & \mw{gyt} & `together' & \mw{y} `to' & yes & c \\
2r & 22 & \mw{teulu} & `retinue' & \mw{y} `his' & no & t \\
2r & 22 & \mw{vaccỽyeit} & `squires' & \mw{e} `his' & yes & m \\
2r & 22 & \mw{wyrda} & `noblemen' & \mw{e} `his' & yes & g \\
2r & 23 & \mw{gerdoryon} & `musicians' & \mw{e} `his' & yes & c \\
2r & 24 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
2r & 25 & \mw{vab} & `son' & \mw{neu} & yes & m \\
2r & 25 & \mw{vab} & `son' & apposition & yes & m \\
2r & 25 & \mw{vyd} & `shall be' & [\mw{a}] & yes & b \\
2r & 18--19 & \mw{pym/thec} & `fifteen' & \mw{ar} & no & p \\
2v & 2 & \mw{wnel} & `does' & \mw{a} & yes & g \\
2v & 3 & \mw{alanas} & `compensation' & \mw{vn} & yes & g \\
2v & 3 & \mw{uyd} & `shall be' & [\mw{a}] & yes & b \\
2v & 8 & \mw{golofyn} & `column' & feminine article & yes & c \\
2v & 11 & \mw{le} & `place' & \mw{oes} & yes & ll \\
2v & 14 & \mw{treul} & `spending' & preposed adjective & no & t \\
2v & 16 & \mw{bieu} & `owns' & [\mw{a}] & yes & p \\
2v & 16 & \mw{gantaỽ} & `with, by him' & petrification & yes & c \\
2v & 18 & \mw{gyscu} & `sleep' & \mw{y} `to' & yes & c \\
2v & 19 & \mw{teir} & `three' & feminine article & no & t \\
2v & 19 & \mw{vessur} & `measure' & \mw{heb} & yes & m \\
2v & 21 & \mw{paỽb} & `everybody' & parenthesis & no & p \\
2v & 22 & \mw{pop} & `every' & \mw{y} `to' & no & p \\
2v & 23 & \mw{dros} & `over' & petrification & yes & t \\
2v & 23 & \mw{gyrcho} & `approaches' & \mw{a} & yes & c \\
2v & 24 & \mw{wlat} & `land' & feminine article & yes & g \\
2v & 25 & \mw{gan} & `with, by' & petrification & yes & c \\
2v & 11--12 & \mw{ỽr/thrychyeit} & `successors' & preposed adjective & yes & g \\
2v & 6--7 & \mw{gyfar/ỽyneb} & `facing' & adverbial clause & yes & c \\
3r & 1 & \mw{dros} & `over' & petrification & yes & t \\
3r & 1 & \mw{teruyn} & `end' & \mw{dros} & no & t \\
3r & 3 & \mw{weryt} & `delivers' & \mw{a} & yes & g \\
3r & 5 & \mw{gyscu} & `sleep' & \mw{y} `to' & yes & c \\
3r & 7 & \mw{parha} & `remains' & \mw{a} & no & p \\
3r & 8 & \mw{gorn} & `horn' & \mw{y} `his' & yes & c \\
3r & 9 & \mw{baraho} & `stays' & \mw{?tra} & yes & p \\
3r & 9 & \mw{gyntaf} & `first' & feminine noun & yes & c \\
3r & 10 & \mw{parha} & `remains' & \mw{a} & no & p \\
3r & 12 & \mw{urỽynha} & `gather rushes' & \mw{y} `to' & yes & b \\
3r & 16 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
3r & 17 & \mw{gyscu} & `sleep' & \mw{y} `to' & yes & c \\
3r & 19 & \mw{ostec} & `silence' & feminine article & yes & g \\
3r & 20 & \mw{gilyd} & `each other' & \mw{e} `his' & yes & c \\
3r & 22 & \mw{gyntaf} & `first' & feminine noun & yes & c \\
3r & 23 & \mw{dan} & `under' & petrification & yes & t \\
3r & 23 & \mw{traet} & `feet' & \mw{dan} & no & t \\
3r & 21-22 & \mw{ga/nhỽyll} & `candle' & feminine article & yes & c \\
3r--3v & 25--1 & \mw{go/lỽyth} & `spleen' & \mw{y} `his' & no & g \\
3v & 1 & \mw{ossotto} & `places' & \mw{pan} & yes & g \\
3v & 2 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
3v & 4 & \mw{gaffo} & `has' & \mw{pan} & yes & c \\
3v & 6 & \mw{gerỽyn} & `tun' & feminine article & yes & c \\
3v & 6 & \mw{ved} & `mead' & feminine noun & yes & m \\
3v & 8 & \mw{ved} & `mead' & feminine noun & yes & m \\
3v & 9 & \mw{ouỽy} & `visit' & \mw{y} `to' & yes & g \\
3v & 9 & \mw{gan} & `with, by' & petrification & yes & c \\
3v & 9 & \mw{ganhat} & `permission' & \mw{gan} & yes & c \\
3v & 10 & \mw{trachefyn} & `again' & adverbial clause & no & t \\
3v & 12 & \mw{vreich} & `arm' & \mw{y} `his' & yes & b \\
3v & 12 & \mw{wyalen} & `rod' & \mw{e} `his' & yes & g \\
3v & 15 & \mw{lety} & `lodging' & \mw{e} `his' & yes & ll \\
3v & 17 & \mw{para} & `remains' & \mw{a} & no & p \\
3v & 17 & \mw{wnel} & `does' & \mw{tra} & yes & g \\
3v & 19 & \mw{pedolo} & `shoes' & \mw{thra} & no & p \\
3v & 21 & \mw{torher} & `is broken' & \mw{a} & no & t \\
3v & 21 & \mw{bynhac} & `-ever' & \mw{Pỽy} & yes & p \\
3v & 22 & \mw{telir} & `is paid' & \mw{a} & no & t \\
3v & 24 & \mw{alanas} & `compensation' & \mw{y} `his' & yes & g \\
3v & 24 & \mw{uelly} & `thus' & petrification & yes & m \\
3v & 16--17 & \mw{gi/lyd} & `each other' & \mw{e} `his' & yes & c \\
3v & 7--8 & \mw{ge/rỽyn} & `tun' & feminine article & yes & c \\
6r & 1 & \mw{laỽ} & `hand' & \mw{uch} & yes & ll \\
6r & 2 & \mw{laỽ} & `hand' & \mw{is} & yes & ll \\
6r & 2 & \mw{teir} & `three' & feminine article & no & t \\
6r & 3 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6r & 5 & \mw{wnel} & `does' & \mw{a} & yes & g \\
6r & 7 & \mw{gyfreith} & `law' & \mw{ỽrth} & yes & c \\
6r & 8 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6r & 9 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6r & 9 & \mw{gynt} & `before' & \mw{yn} & yes & c \\
6r & 10 & \mw{bieu} & `owns' & [\mw{a}] & yes & p \\
6r & 11 & \mw{uuch} & `cattle' & \mw{neu} & yes & b \\
6r & 12 & \mw{dros} & `over' & petrification & yes & t \\
6r & 12 & \mw{vo} & `is' & \mw{pan} & yes & b \\
6r & 12 & \mw{bieu} & `owns' & [\mw{a}] & yes & p \\
6r & 13 & \mw{geidỽ} & `keeps' & \mw{a} & yes & c \\
6r & 16 & \mw{gaffo} & `has' & \mw{pan} & yes & c \\
6r & 16 & \mw{gan} & `with, by' & petrification & yes & c \\
6r & 17 & \mw{barn} & `judgment' & \mw{y} `his' & no & b \\
6r & 17 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6r & 18 & \mw{perthyno} & `relates' & \mw{a} & no & p \\
6r & 20 & \mw{gan} & `with, by' & petrification & yes & c \\
6r & 20 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6r & 21 & \mw{vreint} & `sovereignty' & \mw{y} `his' & yes & b \\
6r & 23 & \mw{ỽr} & `man' & \mw{deu} & yes & g \\
6r & 23 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6r & 24 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6r & 24 & \mw{wnel} & `does' & \mw{a} & yes & g \\
6r & 25 & \mw{ty} & `house' & \mw{e} `his' & no & t \\
6r & 18--19 & \mw{bi/eu} & `owns' & [\mw{a}] & yes & p \\
6v & 1 & \mw{ỽystyl} & `pledge' & \mw{deu} & yes & g \\
6v & 3 & \mw{varnet} & `must judge' & \mw{na} & yes & b \\
6v & 3 & \mw{tauaỽt} & `tongue' & \mw{y} `his' & no & t \\
6v & 3 & \mw{werth} & `worth' & parenthesis & yes & g \\
6v & 5 & \mw{tauaỽt} & `tongue' & \mw{y} `his' & no & t \\
6v & 5 & \mw{werth} & `worth' & parenthesis & yes & g \\
6v & 6 & \mw{kaffel} & `receive' & parenthesis & no & c \\
6v & 7 & \mw{pop} & `every' & \mw{o} & no & p \\
6v & 7 & \mw{talo} & `pays' & \mw{a} & no & t \\
6v & 9 & \mw{teruyner} & `is ended' & \mw{pan} & no & t \\
6v & 10 & \mw{ganhat} & `permission' & \mw{heb} & yes & c \\
6v & 13 & \mw{ỽyppo} & `knows' & \mw{ny} & yes & g \\
6v & 13 & \mw{varnu} & `judge' & object lenition & yes & b \\
6v & 15 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6v & 16 & \mw{pressỽyl} & `presence' & \mw{yn} & no & p \\
6v & 16 & \mw{gan} & `with, by' & petrification & yes & c \\
6v & 16 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
6v & 17 & \mw{gan} & `with, by' & petrification & yes & c \\
6v & 17 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6v & 19 & \mw{peunydyaỽl} & `daily' & adverbial clause & no & p \\
6v & 20 & \mw{varch} & `horse' & \mw{y} `his' & yes & m \\
6v & 20 & \mw{gyweir} & `working order' & \mw{yn} & yes & c \\
6v & 21 & \mw{tir} & `land' & \mw{y} `his' & no & t \\
6v & 21 & \mw{tlysseu} & `jewelry' & preposed adjective & no & t \\
6v & 21 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6v & 21 & \mw{geiff} & `has' & \mw{a} & yes & c \\
6v & 22 & \mw{ỽystler} & `is pledged' & \mw{pan} & yes & g \\
6v & 22 & \mw{gan} & `with, by' & petrification & yes & c \\
6v & 23 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
6v & 23 & \mw{gan} & `with, by' & petrification & yes & c \\
6v & 24 & \mw{gadu} & `keep' & object lenition & yes & c \\
6v & 25 & \mw{gan} & `with, by' & petrification & yes & c \\
6v & 25 & \mw{werth} & `worth' & \mw{ar} & yes & g \\
6v & 24--25 & \mw{gan/taỽ} & `with, by him' & petrification & yes & c \\
7r & 3 & \mw{gadeir} & `chair' & \mw{y} `his' & yes & c \\
7r & 3 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7r & 4 & \mw{pop} & `every' & \mw{o} & no & p \\
7r & 5 & \mw{gan} & `with, by' & petrification & yes & c \\
7r & 6 & \mw{pen} & `head' & \mw{y} `to' & no & p \\
7r & 6 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7r & 8 & \mw{uarn} & `judgment' & \mw{a} & yes & b \\
7r & 9 & \mw{gehyr} & `piece of meat' & \mw{o} & yes & c \\
7r & 11 & \mw{gynheil} & `maintains' & \mw{a} & yes & c \\
7r & 12 & \mw{uyd} & `shall be' & [\mw{a}] & yes & b \\
7r & 14 & \mw{bynhac} & `-ever' & adverbial clause & yes & p \\
7r & 15 & \mw{vynyd} & `mountain' & feminine noun & yes & m \\
7r & 15 & \mw{bỽn} & `bittern' & \mw{neu} & no & b \\
7r & 16 & \mw{rym} & `force' & \mw{o} & yes & g \\
7r & 16 & \mw{wna} & `does' & \mw{a} & yes & g \\
7r & 17 & \mw{varch} & `horse' & \mw{y} `his' & yes & m \\
7r & 18 & \mw{warthafyl} & `stirrup' & \mw{y} `his' & yes & g \\
7r & 20 & \mw{laỽ} & `hand' & \mw{e} `his' & yes & ll \\
7r & 20 & \mw{uỽyt} & `food' & \mw{ar} & yes & b \\
7r & 21 & \mw{beunyd} & `every day' & adverbial clause & yes & p \\
7r & 21 & \mw{gennat} & `messenger' & \mw{y} `his' & yes & c \\
7r & 22 & \mw{teir} & `three' & feminine article & no & t \\
7r & 23 & \mw{gled} & `left' & \mw{ar} & yes & c \\
7r & 24 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7r & 25 & \mw{gan} & `with, by' & petrification & yes & c \\
7r & 25 & \mw{wneuthur} & `do' & \mw{y} `to' & yes & g \\
7v & 3 & \mw{gan} & `with, by' & petrification & yes & c \\
7v & 3 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7v & 4 & \mw{varch} & `horse' & \mw{y} `his' & yes & m \\
7v & 6 & \mw{bieu} & `owns' & [\mw{a}] & yes & p \\
7v & 6 & \mw{gan} & `with, by' & petrification & yes & c \\
7v & 6 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7v & 7 & \mw{tir} & `land' & \mw{ar} & no & t \\
7v & 7 & \mw{bieu} & `owns' & [\mw{a}] & yes & p \\
7v & 7 & \mw{gaffer} & `is had' & \mw{a} & yes & c \\
7v & 8 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7v & 9 & \mw{lety} & `lodging' & \mw{y} `his' & yes & ll \\
7v & 10 & \mw{uo} & `is' & \mw{a} & yes & b \\
7v & 11 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7v & 12 & \mw{weith} & `time' & \mw{vn} & yes & g \\
7v & 12 & \mw{pop} & `every' & adverbial clause & no & p \\
7v & 12 & \mw{tayogeu} & `serfs' & \mw{ar} & no & t \\
7v & 13 & \mw{pop} & `every' & \mw{o} & no & p \\
7v & 14 & \mw{uỽyt} & `food' & \mw{yn} & yes & b \\
7v & 15 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7v & 16 & \mw{lle} & `place' & \mw{y} `his' & no & l \\
7v & 17 & \mw{gantaỽ} & `with, by him' & petrification & yes & c \\
7v & 19 & \mw{wisc} & `clothing' & \mw{y} `his' & yes & g \\
7v & 20 & \mw{lather} & `is killed' & \mw{a} & yes & ll \\
7v & 20 & \mw{bieu} & `owns' & [\mw{a}] & yes & p \\
7v & 20 & \mw{gegin} & `kitchen' & \mw{y} `his' & yes & c \\
7v & 21 & \mw{gyfreith} & `law' & \mw{o} & yes & c \\
7v & 24 & \mw{gan} & `with, by' & petrification & yes & c \\
7v & 24 & \mw{geiff} & `has' & \mw{a} & yes & c \\
7v & 25 & \mw{wneuthur} & `do' & \mw{y} `to' & yes & g \\
8r & 1 & \mw{les} & `benefit' & \mw{ar} & yes & ll \\
8r & 4 & \mw{gỽn} & `hounds' & \mw{y} `his' & yes & c \\
8r & 5 & \mw{gynllyfaneu} & `leashes' & \mw{e} `his' & yes & c \\
8r & 5 & \mw{trayan} & `third' & \mw{e} `his' & no & t \\
8r & 5 & \mw{gyrn} & `horns' & \mw{e} `his' & yes & c \\
8r & 7 & \mw{gantaỽ} & `with, by him' & petrification & yes & c \\
8r & 8 & \mw{uyd} & `shall be' & [\mw{a}] & yes & b \\
8r & 9 & \mw{geiff} & `has' & \mw{a} & yes & c \\
8r & 9 & \mw{gilyd} & `each other' & \mw{y} `his' & yes & c \\
8r & 10 & \mw{ỽr} & `man' & \mw{deu} & yes & g \\
8r & 10 & \mw{gan} & `with, by' & petrification & yes & c \\
8r & 10 & \mw{gynydon} & `huntsmen' & \mw{gan} & yes & c \\
8r & 11 & \mw{gan} & `with, by' & petrification & yes & c \\
8r & 11 & \mw{gynydon} & `huntsmen' & \mw{gan} & yes & c \\
8r & 12 & \mw{trayan} & `third' & \mw{y} `his' & no & t \\
8r & 14 & \mw{gantaỽ} & `with, by him' & petrification & yes & c \\
8r & 16 & \mw{gymryt} & `take' & \mw{y} `to' & yes & c \\
8r & 17 & \mw{gantaỽ} & `with, by him' & petrification & yes & c \\
8r & 18 & \mw{gantaỽ} & `with, by him' & petrification & yes & c \\
8r & 18 & \mw{golofyn} & `column' & feminine article & yes & c \\
8r & 19 & \mw{gan} & `with, by' & petrification & yes & c \\
8r & 20 & \mw{gan} & `with, by' & petrification & yes & c \\
8r & 20 & \mw{gan} & `with, by' & petrification & yes & c \\
8r & 21 & \mw{vrenhines} & `queen' & feminine article & yes & b \\
8r & 21 & \mw{gan} & `with, by' & petrification & yes & c \\
8r & 22 & \mw{vihagel} & (personal name) & feminine noun & yes & m \\
8r & 22 & \mw{pop} & `every' & adverbial clause & no & p \\
8r & 22 & \mw{gan} & `with, by' & petrification & yes & c \\
8r & 22 & \mw{geiff} & `has' & \mw{a} & yes & c \\
8r & 23 & \mw{lety} & `lodging' & \mw{y} `his' & yes & ll \\
8r & 23 & \mw{geiff} & `has' & \mw{a} & yes & c \\
8r & 24 & \mw{bieu} & `owns' & [\mw{a}] & yes & p \\
8r & 11--12 & \mw{tra/yan} & `third' & \mw{o} & no & t \\
8r & 1--2 & \mw{ga/lan} & `calends' & \mw{hyt} & yes & c \\
8r & 18--19 & \mw{gyfarỽyn/eb} & `facing' & feminine noun & yes & c \\
8v & 1 & \mw{Gyt} & `together' & petrification & yes & c \\
8v & 4 & \mw{vei} & `May' & \mw{o} & yes & m \\
8v & 7 & \mw{gan} & `with, by' & petrification & yes & c \\
8v & 7 & \mw{geiff} & `has' & \mw{a} & yes & c \\
8v & 8 & \mw{tygho} & `swears' & \mw{Pan} & no & t \\
8v & 9 & \mw{uỽyn} & `by' & \mw{y} `to' & yes & m \\
8v & 9 & \mw{gỽn} & `hounds' & \mw{y} `his' & yes & c \\
8v & 9 & \mw{gyrn} & `horns' & \mw{y} `his' & yes & c \\
8v & 10 & \mw{geiff} & `has' & \mw{a} & yes & c \\
8v & 11 & \mw{pop} & `every' & \mw{gan} & no & p \\
8v & 11 & \mw{gan} & `with, by' & petrification & yes & c \\
8v & 11 & \mw{geinhaỽc} & `pence' & \mw{ỽyth} & yes & c \\
8v & 12 & \mw{pop} & `every' & \mw{gan} & no & p \\
8v & 12 & \mw{gan} & `with, by' & petrification & yes & c \\
8v & 13 & \mw{teulu} & `retinue' & \mw{gan} & no & t \\
8v & 13 & \mw{gan} & `with, by' & petrification & yes & c \\
8v & 14 & \mw{lu} & `host' & \mw{gan} & yes & ll \\
8v & 14 & \mw{vo} & `is' & \mw{pan} & yes & b \\
8v & 14 & \mw{gan} & `with, by' & petrification & yes & c \\
8v & 14 & \mw{gorn} & `horn' & \mw{y} `his' & yes & c \\
8v & 15 & \mw{geiff} & `has' & \mw{yt} & yes & c \\
8v & 16 & \mw{gan} & `with, by' & petrification & yes & c \\
8v & 17 & \mw{kaffel} & `have' & parenthesis & no & c \\
8v & 18 & \mw{gantaỽ} & `with, by him' & petrification & yes & c \\
8v & 9--10 & \mw{gynlly/uaneu} & `leashes' & \mw{y} `his' & yes & c \\ \bottomrule
\caption{Representation of lenition in \gls{bl} Harley MS 4353 until f.\ 8v l.\ 10}
\label{lenitionharley4353}
\end{mylongtable}

\begin{mylongtable}{@{}rrlllll@{}}
\toprule
\textbf{Page} & \textbf{Line} & \textbf{Word} & \textbf{Translation} & \textbf{Cause of lenition} & \textbf{Represented} & \textbf{Starts w/} \\ \midrule\endhead
17 & 1 & \mw{wenhwyuar } & (personal name) & \mw{y} `to' & yes & g \\
17 & 3 & \mw{gaer } & `fort' & fem. article & yes & c \\
17 & 3 & \mw{gan} & `with' & petrification & yes & c \\
17 & 3 & \mw{wenhwyuar } & (personal name) & \mw{gan} & yes & g \\
17 & 5 & \mw{tebygaf } & `suppose' & \mw{a} & no & t \\
17 & 5 & \mw{wenhwyuar} & (personal name) & \mw{att} & yes & g \\
17 & 6 & \mw{Ereint} & (personal name) & obj.\ lenition & yes & g \\
17 & 6 & \mw{varch} & `horse' & \mw{ar} & yes & m \\
17 & 6 & \mw{vorỽyn } & `maiden' & fem. article & yes & m \\
17 & 7 & \mw{vorỽyn } & `maiden' & fem. article & yes & m \\
17 & 8 & \mw{lieinwisc} & `linen clothing' & \mw{y} `to' & yes & ll \\
17 & 8 & \mw{welỽn } & `saw' & \mw{a} & yes & g \\
17 & 9 & \mw{Gỽenhwyuar} & (personal name) & fem. article & yes & g \\
17 & 9 & \mw{raessaỽ} & `grace' & \mw{y} `his' & yes & g \\
17 & 9 & \mw{wraged } & `women' & prep.\ adjective & yes & g \\
17 & 10 & \mw{oruc } & `did' & \mw{a} & yes & g \\
17 & 10 & \mw{uot } & `to be' & \mw{y} `to' & yes & b \\
17 & 11 & \mw{vorỽyn} & `maiden' & fem. article & yes & m \\
17 & 11 & \mw{wenhwyuar} & (personal name) & \mw{oed} & yes & g \\
17 & 13 & \mw{glotuaỽr} & `famous' & fem. noun & yes & c \\
17 & 14 & \mw{gyn} & `before' & \mw{yn} & yes & c \\
17 & 14 & \mw{peri} & `cause' & parenthesis & no & p \\
17 & 14 & \mw{talho} & `pay' & \mw{a} & no & t \\
17 & 14 & \mw{ualchet} & `as proud' & \mw{gyn} & yes & b \\
17 & 15 & \mw{buchỽn} & `wished' & \mw{ry} & yes & p \\
17 & 15 & \mw{peri } & `cause' & obj.\ lenition & no & p \\
17 & 16 & \mw{uorỽyn } & `maiden' & fem. article & yes & m \\
17 & 17 & \mw{Gỽenhwyuar} & (personal name) & fem. article & no & g \\
17 & 18 & \mw{vot} & `to be' & parenthesis & yes & b \\
17 & 19 & \mw{oruc } & `did' & \mw{a} & yes & g \\
17 & 19 & \mw{orugant} & `did' & \mw{a} & yes & g \\
17 & 22 & \mw{genhyt} & `with you' & petrification & yes & c \\
17 & 22 & \mw{lỽydyanus } & `succesful' & fem. noun & yes & ll \\
17 & 25 & \mw{orffei} & `ended' & \mw{yny} & yes & g \\
17 & 25 & \mw{pỽy } & `who' & obj.\ lenition & no & p \\
17 & 25 & \mw{vei} & `would be' & [\mw{a}] & yes & b \\
17 & 25 & \mw{ỽypỽn } & `knew' & \mw{hyny} & yes & g \\
17 & 26 & \mw{gigleu } & `heard' & \mw{a} & yes & c \\
17 & 26 & \mw{uorỽyn } & `maiden' & fem. article & yes & m \\
17 & 26 & \mw{ỽr} & `man' & voc.\ \mw{a} & yes & g \\
17 & 27 & \mw{gyt } & `with' & petrification & yes & c \\
17 & 28 & \mw{getymdeithon} & `companions' & \mw{e} `his' & yes & c \\
17 & 28 & \mw{uu } & `was' & [\mw{a}] & yes & b \\
17 & 28 & \mw{vorỽyn} & `maiden' & fem. article & yes & m \\
17 & 28 & \mw{welet } & `see' & \mw{y} `to' & yes & g \\
18 & 1 & \mw{gan} & `with' & petrification & yes & c \\
18 & 1 & \mw{paỽb} & `everyone' & \mw{gan} & no & p \\
18 & 1 & \mw{pei} & `if … were' & \mw{ky/hyttrei} & no & p \\
18 & 2 & \mw{vorỽyn } & `maiden' & fem. article & yes & m \\
18 & 2 & \mw{welsynt } & `saw' & \mw{na} & yes & g \\
18 & 3 & \mw{uu } & `was' & \mw{a} & yes & b \\
18 & 3 & \mw{vorỽyn} & `maiden' & parenthesis & yes & m \\
18 & 3 & \mw{ỽympach } & `grander' & fem. noun & yes & g \\
18 & 4 & \mw{ereint} & (personal name) & \mw{y} `to' & yes & g \\
18 & 4 & \mw{wneit } & `was done' & \mw{a} & yes & g \\
18 & 5 & \mw{vorỽyn} & `maiden' & fem. article & yes & m \\
18 & 5 & \mw{wiscoed } & `clothing' & prep.\ adjective & yes & g \\
18 & 6 & \mw{uorỽyn} & `maiden' & fem. article & yes & m \\
18 & 6 & \mw{uorỽyn } & `maiden' & fem. article & yes & m \\
18 & 6 & \mw{welhei } & `saw' & \mw{a} & yes & g \\
18 & 7 & \mw{olỽc } & `sight' & \mw{-ei} & yes & g \\
18 & 7 & \mw{wedeidlỽys } & `handsomeness' & fem. noun & yes & g \\
18 & 7 & \mw{welei } & `saw' & \mw{a} & yes & g \\
18 & 8 & \mw{gerdeu } & `song' & \mw{trỽy } & yes & c \\
18 & 8 & \mw{treulassant } & `spend' & \mw{a} & no & t \\
18 & 9 & \mw{wirodeu } & `strong drinks' & prep.\ adjective & yes & g \\
18 & 10 & \mw{ganthunt } & `with them' & petrification & yes & c \\
18 & 10 & \mw{uu } & `was' & \mw{phan} & yes & b \\
18 & 10 & \mw{vynet } & `go' & parenthesis & yes & m \\
18 & 11 & \mw{wely } & `bed' & existential verb & yes & g \\
18 & 13 & \mw{gyntaf } & `first' & fem. noun & yes & c \\
18 & 13 & \mw{gyt} & `with' & \mw{y} `to' & yes & c \\
18 & 14 & \mw{dros } & `over' & petrification & yes & t \\
18 & 14 & \mw{ereint} & (personal name) & \mw{dros} & yes & g \\
18 & 15 & \mw{uorỽyn } & `maiden' & fem. article & yes & m \\
18 & 16 & \mw{vorỽyn} & `maiden' & \mw{vn } & yes & m \\
18 & 16 & \mw{wyr } & `men' & \mw{o} & yes & g \\
18 & 18 & \mw{pen } & `head' & \mw{am} & no & p \\
18 & 19 & \mw{le} & `place' & \mw{llyma } & yes & ll \\
18 & 19 & \mw{verch } & `daughter' & apposition & yes & m \\
18 & 20 & \mw{uorỽyn } & `maiden' & fem. article & yes & m \\
18 & 21 & \mw{garyat} & `love' & \mw{o} & yes & c \\
18 & 22 & \mw{gan} & `with' & petrification & yes & c \\
18 & 22 & \mw{paỽb } & `everyone' & \mw{gan} & no & p \\
18 & 22 & \mw{uu } & `was' & [\mw{a}] & yes & b \\
18 & 23 & \mw{wnaethpỽyt } & `is done' & \mw{a} & yes & g \\
18 & 25 & \mw{oruc } & `did' & \mw{a} & yes & g \\
18 & 25 & \mw{uỽy } & `more' & \mw{yn} & yes & m \\
18 & 27 & \mw{pop } & `every' & \mw{o} & no & p \\
18 & 28 & \mw{glot } & `fame' & \mw{y} `his' & yes & c \\
18 & 11 & \mw{gyscu } & `sleep' & \mw{y} `to' & yes & c \\
18 & 19 & \mw{Ereint} & (personal name) & \mw{-ei} & yes & g \\
18 & 20 & \mw{glotuoraf} & `praise' & fem. noun & yes & c \\
18 & 3--4 & \mw{vor/ỽyn } & `maiden' & fem. article & yes & m \\
18 & 4--5 & \mw{wnaeth/pỽyt} & `is done' & \mw{a} & yes & g \\
18 & 6--7 & \mw{wisc } & `clothing' & fem. article & yes & g \\
19 & 1 & \mw{gỽeith } & `time' & prep.\ adjective & no & g \\
19 & 1 & \mw{traỽs } & `across' & \mw{ar} & no & t \\
19 & 5 & \mw{le } & `place' & \mw{py} & yes & ll \\
19 & 5 & \mw{py } & `which' & \mw{o} & no & p \\
19 & 6 & \mw{gan} & `with' & petrification & yes & c \\
19 & 6 & \mw{gernyỽ} & (place name) & \mw{o} & yes & c \\
19 & 7 & \mw{gantaỽ} & `with him' & petrification & yes & c \\
19 & 8 & \mw{val } & `like' & petrification & yes & m \\
19 & 9 & \mw{venegi } & `express' & \mw{y} `to' & yes & m \\
19 & 9 & \mw{vot } & `to be' & \mw{y} `his' & yes & b \\
19 & 10 & \mw{kyttiryogyon } & `compatriots' & \mw{e} `his' & no & c \\
19 & 10 & \mw{ỽybot } & `to know' & \mw{o} & yes & g \\
19 & 11 & \mw{tir } & `land' & \mw{y} `his' & no & t \\
19 & 12 & \mw{vab } & `son' & \mw{y} `his' & yes & m \\
19 & 13 & \mw{gadỽ } & `keep' & \mw{y} `to' & yes & c \\
19 & 13 & \mw{gyfoeth} & `riches' & \mw{y} `his' & yes & c \\
19 & 13 & \mw{teruyneu} & `borders' & \mw{y} `his' & no & t \\
19 & 14 & \mw{well } & `better' & \mw{yn} & yes & g \\
19 & 15 & \mw{teruyneu} & `borders' & \mw{y} `his' & no & t \\
19 & 19 & \mw{uelly } & `thus' & petrification & yes & m \\
19 & 20 & \mw{gantaỽ } & `with him' & petrification & yes & c \\
19 & 21 & \mw{gantaỽ } & `with him' & petrification & yes & c \\
19 & 22 & \mw{gadỽ } & `keep' & \mw{y} `to' & yes & c \\
19 & 22 & \mw{gyfoeth } & `riches' & \mw{y} `his' & yes & c \\
19 & 22 & \mw{gynhal } & `maintain' & \mw{y} `to' & yes & c \\
19 & 22 & \mw{teruyneu} & `borders' & \mw{y} `his' & no & t \\
19 & 23 & \mw{lei} & `less' & adv.\ degr. & yes & ll \\
19 & 23 & \mw{tat} & `father' & \mw{y} `his' & no & t \\
19 & 24 & \mw{vorynyon } & `maidens' & prep.\ adjective & yes & m \\
19 & 24 & \mw{wraged } & `women' & prep.\ adjective & yes & g \\
19 & 25 & \mw{treulassant } & `spend' & \mw{a} & no & t \\
19 & 26 & \mw{ereint} & (personal name) & \mw{y} `to' & yes & g \\
19 & 26 & \mw{pop } & `every' & \mw{o} & no & p \\
19 & 26 & \mw{venegis } & `expressed' & \mw{a} & yes & m \\
19 & 27 & \mw{gernyỽ } & (place name) & \mw{o} & yes & c \\
19 & 28 & \mw{les } & `benefit' & \mw{o} & yes & ll \\
19 & 11 & \mw{gyfoeth} & `riches' & \mw{e} `his' & yes & c \\
19 & 18--19 & \mw{gef/fỽch} & `have' & \mw{a} & yes & c \\
19 & 19 & \mw{oruc} & `did' & \mw{a} & yes & g \\
19 & 22--23 & \mw{all/ei} & `could' & \mw{kany} & yes & g\\ \bottomrule
\caption{Representation of lenition in pp.\ 17--19 of \mw{Geraint ac Enid} in \gls{nlw} MS.\ Peniarth 6.}
\label{geraintenid}
\end{mylongtable}
