\message{ !name(../main.tex)}\documentclass[
% draft,
a4paper,
12pt,
twoside,
openright
]{memoir}
\chapterstyle{thatcher}

\setlrmargins{*}{*}{1.618033}
\setlength{\textheight}{1.618033\textwidth}
\setulmargins{*}{*}{1.618033}

\pagestyle{headings}
\nouppercaseheads

\setsecnumdepth{subsection}
\renewcommand\labelenumi{(\theenumi)}

\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{positioning,calc}
\usepackage{forest}
\usepackage{pgfplots}
% \usepackage{pgfgantt}
\usepackage{rotating}
\usepackage{xstring}
\usepackage{lscape}

\usepackage{fontspec}
\usepackage{polyglossia}
    \setmainlanguage[variant=british]{english}
    \setotherlanguage{welsh}
    \setotherlanguage{french}
    \setotherlanguage{latin}
    \setotherlanguage[variant=ancient]{greek}
    \setotherlanguage{irish}
    \setotherlanguage{breton}
    \input{fixwelsh}

\setmainfont[
            RawFeature={+mgrk;+liga;+clig;+ccmp;+tlig;+rlig;+calt;+kern;+onum;+ss01;+falt},
            BoldFont=Junicode-Bold.ttf,
            ItalicFont=Junicode-Italic.ttf,
            BoldItalicFont=Junicode-BoldItalic.ttf,
            ]{Junicode.ttf}
\usepackage{biolinum}
\setmonofont{FantasqueSansMono-Regular.ttf}[%
            Scale=MatchLowercase,
            BoldFont=FantasqueSansMono-Bold.ttf,
            ItalicFont=FantasqueSansMono-Italic.ttf,
            BoldItalicFont=FantasqueSansMono-BoldItalic.ttf]

\usepackage{ellipsis}
% \usepackage[config=mt-Junicode]{microtype}
\usepackage{xspace}

\usepackage[
            language=british,
            backend=biber,
            style=authoryear,
            isbn=false,
            doi=false,
            ]{biblatex}
            \addbibresource{latbib.bib}
            \renewcommand{\bibnamedash}{\rule[.5ex]{1.5em}{.1ex}\space}
            \AtEveryBibitem{\clearlist{language}}
        
\usepackage{longtable}
\usepackage{multirow}
\usepackage{tablefootnote}
\usepackage{booktabs}
\newcolumntype{d}{>{\addfontfeature{RawFeature={+lnum;-onum}}}r}
\newcolumntype{D}{>{\addfontfeature{RawFeature={+lnum;-onum}}}l}
\usepackage{collcell}
\newcolumntype{w}{>{\collectcell\mw}l<{\endcollectcell}}
\newcolumntype{q}{>{\collectcell\noparquotes}l<{\endcollectcell}}
\newcolumntype{a}{>{\collectcell\acrshort}l<{\endcollectcell}}
\newcolumntype{I}{>{\itshape}l}
% \newcolumntype{L}{>{/}c<{/}}
% \newcolumntype{T}{>{{[}}c<{{]}}}
\newcommand{\tchhh}[1]{\multicolumn{3}{c}{#1}}
\newcommand{\tchh}[1]{\multicolumn{2}{c}{#1}}
\newcommand{\tch}[1]{\multicolumn{1}{c}{#1}}
\usepackage{subfig}

\newcommand{\noparquotes}[1]{%
  \IfBeginWith{#1}{(}%
  {%
    \IfEndWith{#1}{)}
    {%
      #1
    }%
    {`#1'}%end with ) 
  }%If begins with (
  {`#1'}%begin with (
}

\usepackage{csquotes}

\usepackage{hyphenat}
\AtBeginDocument{\lefthyphenmin=4\righthyphenmin=4}
\hyphenpenalty=1300

\usepackage{float}
\usepackage[section]{placeins} %Prevents floats from crossing section boundaries
\usepackage{calc}
\usepackage{etoolbox}

\usepackage[color=orange!40!white!40!yellow!20,linecolor=gray,textsize=footnotesize]{todonotes}
\usepackage[normalem]{ulem} %underlining

\usepackage{graphicx} % For images
\graphicspath{ {images/} }

\ifdraftdoc
\usepackage{datetime}
\newcommand*{\draftfoottext}{\textwelsh[date=long]{\texttt{\tiny Drafft: \today\ @\currenttime}}}
  \makeevenfoot{plain}{\texttt{\tiny Paulus van Sluis}}{\thepage}{\draftfoottext}
  \makeoddfoot{plain}{\draftfoottext}{\thepage}{\texttt{\tiny Paulus van Sluis}}
  \makeevenfoot{headings}{}{}{\draftfoottext}
  \makeoddfoot{headings}{\draftfoottext}{}{}
\fi

\DisemulatePackage{setspace}%Use when class=memoir
\usepackage{setspace}
\setstretch{1.5} %Enable to receive one-half spacing
\newcommand{\onesp}[1]{\begin{spacing}{1}#1\end{spacing}}

\usepackage[marginal]{footmisc}

%% NO DOT: vs
\makeatletter
\newcommand*{\ie}{\@ifnextchar{.}{i.e}{i.e.\@\xspace}}
\newcommand*{\eg}{\@ifnextchar{.}{e.g}{e.g.\@\xspace}}
\newcommand*{\etc}{\@ifnextchar{.}{etc}{etc.\@\xspace}}
\newcommand*{\cf}{\@ifnextchar{.}{cf}{cf.\@\xspace}}
\makeatother

\input{langcom}

\newcommand{\al}[1]{\uline{#1}}

\newcommand{\oes}{\mw{oes}}
\newcommand{\ei}{\mw{\mbox{-ei}}}
\newcommand{\graph}[1]{\textnormal{\textup{⟨}}#1\textnormal{\textup{⟩}}}
\newcommand{\abbr}[1]{\textup{[#1]}}

\newcommand{\gd}{g̊}
\newcommand{\gwd}{g̊ʷ}
\newcommand{\bd}{b̥}
\newcommand{\dd}{d̥}
\newcommand{\cw}{w̯}
\newcommand{\cu}{u̯}
\newcommand{\ci}{i̯}
\newcommand{\rh}{r̥}
\newcommand{\Rh}{R̥}
\newcommand{\zero}{∅}
%Ỻ ỻ Ỽ ỽ Ẏ ẏ

\newcommand{\lT}{\gls{l}\gls{T}}
\newcommand{\xD}{\gls{x}\gls{D}}
\newcommand{\xT}{\gls{x}\gls{T}}
\newcommand{\lD}{\gls{l}\gls{D}}

\newcounter{equationsave}

\newcommand{\mwcc}[4][]{\begin{mwl}\mwc[#1]{#2}{#3}{#4}\end{mwl}}

\newcommand{\mwc}[4][]{%
  \ifstrempty{#1}{\item{\mw{#3} \mbox{(#2)}\\*\textquoteleft #4\textquoteright}}
  {\item{\label{#1}\mw{#3} \mbox{(#2)}\\*\textquoteleft #4\textquoteright}}
}
\newcommand{\langc}[4][]{%
  \ifstrempty{#1}{\item{#3 \mbox{(#2)}\\*\textquoteleft #4\textquoteright}}
  {\item{\label{#1}#3 \mbox{(#2)}\\*\textquoteleft #4\textquoteright}}
}

    
\newenvironment{mwl}{%
    \begin{spacing}{1}%
    \begin{list}{\theequation}{%
        \setcounter{equationsave}{\arabic{equation}}%
        \usecounter{equation}%
        \setcounter{equation}{\arabic{equationsave}}%
        \setlength{\labelsep}{\parindent+\marginparsep}%
        \setlength{\leftmargin}{\parindent}%
        \setlength\labelwidth{3em}%
    }\small}%
    {\end{list}\end{spacing}}

\newcommand{\tqt}[3]{%
  \begin{displayquote}%
    {\small\onesp{#1~\autocite[#3]{#2}}}%
    \end{displayquote}%
  }%quote, reference, page

\newenvironment{mylongtable}[1]%
    {\begin{spacing}{1}\begin{longtable}[c]{#1}}%
    {\end{longtable}\end{spacing}}%

\usepackage[nointegrals]{wasysym}
\newcommand{\TRUE}{\CheckedBox}
\newcommand{\FALSE}{\Square}

% \input{misc/wordcount} Excludes parts; Activate to count words without examples, footnotes and page numbers.

\input{gloss}

% \usepackage[series={},nocritical,noend,noeledsec,nofamiliar,noledgroup]{reledmac}
% \usepackage{reledpar} %%For parallel texts

\begin{document}

\message{ !name(2phon/phonintro.tex) !offset(-194) }
\section{A two-stage development of lenition}
\label{sec:two-stage-devel}
A typical historical grammar describing lenition is \textcite{Mor_Welsh13}, who describes it as follows:
\tqt{Brit.\ and Lat. \textit{p, t, k, b, d, g, m} between vowels became \textit{b, d, g, f, δ, ᵹ, f} respectively in W.}{Mor_Welsh13}{§~103}
This description summarises the totality of changes between \gls{pc} and \gls{mow} correctly, but does not break it up into intermediate stages. 
Yet when he says that present-day Brittonic languages /p t k/ lenite to /b d ɡ/, he says two things: the first is that lenition applied to voiceless stops at some point in the history of Brittonic; the second is that the phonemes to which these voiceless stops lenited merged with the pre-existing phonemes /b d ɡ/. These two developments are often taken together.  Not breaking these two developments up creates a few problems.

One obvious problem is Irish. If the two Brittonic developments occurred at the same time, then lenition in Goidelic and Brittonic cannot be described as a single development, because Goidelic voiceless stops lenite into /f θ x/, and it is hard to derive the Goidelic  voiceless fricatives from the Brittonic voiced stops, or vice versa. Another problem is Brittonic-internal. If voiceless stops after lenition came to equal unlenited voiced stops immediately, then would these not have to be lenited into lenited voiced stops in turn? 

This issue was recognised  by \textcite{loth_les_1892}, who proposed that voiceless stops underwent lenition at a later date than other consonants:
\tqt{\textfrench{Les explosives sonores \textit{b, d, g}, entre deux voyelles ont dû \textit{commencer} leur mouvement vers les spirantes correspondantes, avant que les explosives sourdes, \textit{p, t, c} ne fussent devenues \textit{b, d, g}; autrement, celles-ci auraient eu le même sort qu'elles. Si le latin \textit{opera} avait donné \textit{obera} au moment où \textit{labore} était encore \textit{labure}, le \textit{b} d'\textit{ober} eût été traité comme celui de \textit{labur}; c'est-à-dire fût devenu \textit{v}; on aurait aujourd'hui \textit{over, lavur} et non \textit{ober, lavur}. Des trois explosives sonores, \textit{g} paraît la première être devenue spirante.}\footnote{The voiced stops \textit{b, d, g} between two vowels should have \emph{started} their movement towards their corresponding spirants before the voiceless stops became \textit{b, d, g}; otherwise, these ones would have had the same fate as those. If Latin \textit{opera} had given \textit{obera} at the moment when \textit{labore} was still \textit{labure}, the \textit{b} of \textit{ober} would have been treated as the one of \textit{labur}; \ie it would have become \textit{v}; today we would have \textit{over, lavur} and not \textit{ober, lavur}. Of the three voiced stops, \textit{g} seems to be the first to have become a spirant.}}{loth_les_1892}{87}
Loth's explanation explains why we do not have anything like /p/ > /b/ > /v/, but it has its drawbacks. Lenition may be described as a single process whereby intervocalic consonants become more sonorous, so it would be uneconomical to suggest two separate events of lenition. Incidentally, his account solves the issue of Goidelic lenition:  lenited voiceless stops to voiceless spirants could be just one more independent development of lenition. Yet by now we have to assume three independent events to account for all the lenition in Goidelic and Brittonic.

Alternatively, Loth's explanation could be criticised for not considering chain shifts: it would be economical to assume that intervocalic /p/ and /b/ would shift at the same time in a push chain or a drag chain. Accounting for this possibility, Loth only demonstrates  that lenition of voiceless stops did not  occur \emph{before} other types of lenition, but not necessarily that it must have occurred \emph{afterwards}. Considering Brittonic lenition a chain shift could work for Brittonic itself, but it would force one to consider Goidelic lenition a wholly separate affair.

\Textcite[162]{Foer_Flussname41} and \Textcite{sims-williams_dating_1990} also argue for two separate stages of lenition, with /p t k/ undergoing lenition at a later date than other consonants.
\tqt{I shall argue that `lenition' — the conventional name for the spirantization of [b d g m] > [β δ γ μ] and voicing of [p t k] > [b d g] — occurred in two distinct stages: (1) spirantization, then (2) voicing […]. This will seem heretical to Brittonicists, who instinctively regard [t] > [d], etc., and [d] > [δ], etc., as a single phenomenon because they constitute the morphophonemic alternation of `lenition' or `soft mutation'.}{sims-williams_dating_1990}{221}
He also applies Loth's line of reasoning on how /p t k/ cannot have preceded other consonants because that would have led to these consonants being lenited twice over~\autocite[232]{sims-williams_dating_1990}. This shows that he silently equates lenition of voiceless stops with the merger of lenited voiceless stops with radical voiced stops. Other elements of his argument rest on early loans from Brittonic into Goidelic, but those are rejected by \textcite{isaac_chronology_2004}, because Sims-Williams conflates the roles  phonetics and phonology play in the process of loaning.




\message{ !name(../main.tex) !offset(-24) }

\end{document}

%%% Local Variables:
%%% coding: utf-8
%%% mode: latex
%%% TeX-master: t
%%% End:
