\chapter*{Acknowledgements}%
\label{cha:ack}

First and foremost, I would like to thank my supervisors John Koch and Dafydd Johnston. Their guidance and expertise have been invaluable in every step along the way to completion, as has their genuine interest in my work. I am in their debt for the many hours of discussion and copious amounts of ink left on drafts.

My thanks also go to all my colleagues at the Centre for Advanced Welsh and Celtic Studies. I thank Nia Davies and Angharad Elias for helping with all the practical issues involved in thesis writing  and moving overseas, Marion Loeffler and Elizabeth Edwards for organising all the  training events. My gratitude also goes out to my fellow students, particularly Jeanne Mehan, who helped improve my writing throughout my studies.

Several people commented on early drafts of my work. These include Brynley Roberts, whose comments  immensely improved my chapter on the \mw[]{Brut y Brenhinedd}, Ifan Johnston, who commented on the quantitative aspects of \mw[]{Buchedd Dewi}, and Ann Parry Owen, who kindly commented on my work on the cynghanedd. I am also grateful for all the  discussions with  Ian Hughes, Pavel Iosad, Anders Jørgensen, Gary German, and many others.

I also thank my funders: the University of Wales and the Arts and Humanities Research Council Centre for Doctoral Training in Celtic Languages. 

I would like to thank all the speakers of Welsh and Breton. It is not given that these languages should still exist in today's day and age, yet the thesis could not be written without the living Welsh and Breton that I learned.

Tot slot gaat mijn dank uit naar mijn geliefde Femke. Met haar mocht ik dit avontuur in Wales meemaken. Zij was altijd mijn steun en toeverlaat en bewaakte streng de tijd. 

% Thanks to:
% \begin{itemize}
% \item supervisors
%   \begin{itemize}
% \item John Koch
% \item Dafydd Johnston
% \end{itemize}
% \item CAWCS in general
% \item UoW and AHRC CDT in Celtic languages for funding
% \item People who read/commented
%   \begin{itemize}
%   \item Jeanne Mehan,
%   \item Brynley Roberts
%   \item Mab Dafydd --- enw?
%   \item APO
%   \end{itemize}
% \item People who gave articles/insights/skills developed
%   \begin{itemize}
%   \item Pavel Iosad
%   \item Paul Russell?
%   \item who else?
%   \end{itemize}
% \item Femke --- obviously
  
% \end{itemize}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:
